BOLD Parameter Estimation Using Particle Filters

### 2010-07-22 doc/thesis/results.tex
fixed the histograms for slicesim

\begin{table}[t]
\centering
\begin{tabular}{|c | c | c | c | c | c | c | c | c |}
\hline 
$\tau_0$ & $\alpha$ & $E_0$    & $V_0$    & $\tau_s$ & $\tau_f$ & $\epsilon$  &  $\sqrt{MSR}$   \\
\hline 
1.0324 & 0.33211 & 0.34058 & 0.03012 & 1.40665 & 2.52079 & 0.5311 & todo \\
 0.98189 & 0.33047 & 0.3386 & 0.03014 & 1.45707 & 2.47232 & 0.45049 & 0.00158551 \\
 1.0429 & 0.33224 & 0.34124 & 0.02946 & 1.4618 & 2.49245 & 0.43012 &  0.00165055 \\
 1.02054 & 0.3321 & 0.33484 & 0.02586 & 1.45848 & 2.48741 & 0.4193 &  0.00150604 \\
 1.0565 & 0.33405 & 0.33758 & 0.02791 & 1.43784 & 2.52545 & 0.47517 & 0.00151513 \\
 1.01867 & 0.33528 & 0.33918 & 0.02782 & 1.48345 & 2.49605 & 0.44209 &0.00155645 \\
 1.051 & 0.33038 & 0.33837 & 0.02985 & 1.47651 & 2.48621 & 0.42719 &  0.00158526 \\
 1.00281 & 0.32929 & 0.33988 & 0.0298 & 1.43519 & 2.49256 & 0.48899 & 0.00164299 \\
 1.00893 & 0.33273 & 0.33982 & 0.0289 & 1.42903 & 2.49754 & 0.45688 & 0.0016793  \\
 1.01289 & 0.33275 & 0.3376 & 0.02997 & 1.41188 & 2.49881 & 0.50628 & 0.00182622 \\
 1.10247 & 0.33371 & 0.3419 & 0.02939 & 1.43774 & 2.53384 & 0.44079 & 0.00195009 \\
\hline                                                                
1.03009 & 0.33228 & 0.33905 & 0.02902 & 1.44506 & 2.50031 & 0.46076 & 0.00165129 \\
\hline 
\end{tabular}
\caption{Estimated Parameters on 11 different runs with low noise and no signal present.}
\label{tab:NoiseOnlyResults} 
\end{table}

@@

\subsection{Single Voxel Review}
\begin{table}[t]
\centering
\begin{tabular}{|c | c | c | c | c | c | c | c | c |}
\hline 
& \multicolumn{4}{|c|}{Signal} & \multicolumn{4}{|c|}{No Signal}\\
\hline
& \multicolumn{2}{|c|}{Low Noise} & \multicolumn{2}{|c|}{High Noise} 
& \multicolumn{2}{|c|}{$\sigma_y = .001, \sigma_x = .0005$} 
& \multicolumn{2}{|c|}{$\sigma_y = .01, \sigma_x = .005$}\\
\hline 
& $M.I.$ & N. Res. &
  $M.I.$ & N. Res. &
  $M.I.$ & N. Res. &
  $M.I.$ & N. Res. \\
\hline       
\hline       
1 & 0.86687 &0.47801 &0.09077 &1.03894 &0.06326  &  1.29501 &0.03024   &1.33641 \\
2 & 0.93975 &0.53177 &0.13767 &0.95165 &-0.01075  & 1.30175 &-0.02677  &1.33667 \\
3 & 0.82382 &0.5458  &0.13505 &0.99539 &0.02345  &  1.26287 &-0.0111   &1.15957 \\
4 & 0.94661 &0.49824 &0.04341 &1.16129 &-0.00906  & 1.43196 &0.00147   &1.09988 \\
5 & 0.94281 &0.46805 &0.13718 &1.03972 &0.00663  &  1.25664 &-0.00204  &1.20107 \\
6 & 0.92539 &0.459   &0.12337 &1.00214 &-0.00816  & 1.2708 & 0.01775   &1.04589 \\
7 & 0.98892 &0.46096 &0.15381 &1.08847 &0.02664  &  1.15441 &0.03163   &1.20543 \\
8 & 0.98796 &0.51838 &0.11325 &1.05962 &0.03285  &  1.27456 &0.01951   &1.1225 \\
9 & 0.8804  &0.5253  &0.09669 &1.0157  &0.01628  &  1.32024 &0.01039   &1.08637 \\
10 & 0.88721 &0.49211 &0.18339 &1.18996 &0.00407  &  1.34456 &0.00508   &1.22135 \\
11 & 0.96644 &0.49092 &0.10949 &0.95368 &0.03323  &  1.32522 &-0.01284  &1.11737 \\
\hline                                                                           
min  & 0.92329 &0.49714 &0.12037 &1.04514 &0.01622  &  1.29437 &0.00576   &1.17568 \\
\hline                                                                           
max & 0.96644 &0.459   &0.10949 &0.95165 &-0.01075  & 1.15441 &-0.01284  &1.04589 \\
\hline                                                                           
mean  & 0.86687 &0.5458  &0.09077 &1.18996 &0.06326  &  1.43196 &0.03024   &1.33667 \\
\hline
\end{tabular}
\caption{Mutual Information and the normalized $\sqrt{MSE}$, 
    for signal/noise configurations. TODO FIX MIN/MAX/MEAN}
\label{tab:SignleVoxelActivationComparison} 
\end{table}

The first step in determining the validity of a model is to provide some order of quality
to rate the results by. Unfortunately, because of the variability in the signal levels the
raw $\sqrt{MSR}$ cannot perform this task. As demonstrated by \autoref{sec:PureNoiseLowMag},
a low residual does not necessarily indicate a good fit. 
Therefore, instead I  normalized the signal based on a value proportional to
the magnitude of the signal. Considering the tendency of FMRI noise to have large unexplainable
peaks and troughs, rather than using MAX and MIN values to estimate the scale of the 
signal, I used a robust estimator of 
scale; the Median-Absolute-Deviation (MAD). This is an estimator of the standard
deviation, and thus a good estimator of the scale of the input signal. 
The normalized residual values are 
shown in \autoref{tab:SignleVoxelActivationComparison}. A second potential method of 
gauging performance is mutual information. Mutual information is a method of measuring
the interdependence of two random variables. If two signals are truly independent,
then the mutual information will be zero. Although ideally suited to Bernoulli 
random variables where there are a finite number of options, by using histograms
it is possible to derive a joint distribution of two signals. The algorithm for 
mutual information is based on that joint distribution:
\begin{equation}
\sum_{x,y} p(x,y) \log_2\left(\frac{p(x,y)}{p(x)p(y)}\left)
\end{equation}
Unfortunately the number of bins causing bias in the output, thus to correct for this,
I subtracted the estimated bias:
\begin{equation}
\text{bias} = \frac{N_{bins}}{2Nlog(2)}
\end{equation}
where $N$ is the number of samples and $N_{bins}$ is the number of bins. For all
the mutual information estimates in this work 6 bins are used for the marginal
distribution of each signal. Additionally, throughout log base 2 will be used.
This leads to 36 total bins in the joint, so the bias is: 
\begin{equation}
\text{bias} = \frac{18}{N}
\end{equation}
After calculating the empirical mutual information, I subtracted the bias from
the estimated mutual information. Note that this can result in negative mutual
information, which should not technically be possible; so any negative mutual information
was taken as 0.

@@

\begin{figure} %bottom left
\centering
\includegraphics[clip=true,trim=2.5cm 2cm 2cm 1cm,width=15cm]{images/slicesim_hist1}
\caption{Histogram of estimated parameters in section 1 with $\sqrt{MSR}$ less than $.85$.}
\label{fig:slicesim_hist1}
\end{figure}

\begin{figure} %top left
\centering
\includegraphics[clip=true,trim=2.5cm 2cm 2cm 1cm,width=15cm]{images/slicesim_hist2}
\caption{Histogram of estimated parameters in section 2 with $\sqrt{MSR}$ less than $.85$.}
\label{fig:slicesim_hist2}
\end{figure}

\begin{figure} %top right
\centering
\includegraphics[clip=true,trim=2.5cm 2cm 2cm 1cm,width=15cm]{images/slicesim_hist3}
\caption{Histogram of estimated parameters in section 3 with  $\sqrt{MSR}$ less than $.85$.}
\label{fig:slicesim_hist3}
\end{figure}
todo regenerate histograms

>>>>

\begin{table}[t]
\centering
\begin{tabular}{|c | c | c | c | c | c | c | c | c |}
\hline 
$\tau_0$ & $\alpha$ & $E_0$    & $V_0$    & $\tau_s$ & $\tau_f$ & $\epsilon$  &  $\sqrt{MSR}$   \\
\hline 
1.0324 & 0.33211 & 0.34058 & 0.03012 & 1.40665 & 2.52079 & 0.5311 &   0.00167  \\
 0.98189 & 0.33047 & 0.3386 & 0.03014 & 1.45707 & 2.47232 & 0.45049 & 0.00159   \\
 1.0429 & 0.33224 & 0.34124 & 0.02946 & 1.4618 & 2.49245 & 0.43012 &  0.00165   \\
 1.02054 & 0.3321 & 0.33484 & 0.02586 & 1.45848 & 2.48741 & 0.4193 &  0.00151   \\
 1.0565 & 0.33405 & 0.33758 & 0.02791 & 1.43784 & 2.52545 & 0.47517 & 0.00152   \\
 1.01867 & 0.33528 & 0.33918 & 0.02782 & 1.48345 & 2.49605 & 0.44209 &0.00156   \\
 1.051 & 0.33038 & 0.33837 & 0.02985 & 1.47651 & 2.48621 & 0.42719 &  0.00159   \\
 1.00281 & 0.32929 & 0.33988 & 0.0298 & 1.43519 & 2.49256 & 0.48899 & 0.00164   \\
 1.00893 & 0.33273 & 0.33982 & 0.0289 & 1.42903 & 2.49754 & 0.45688 & 0.00168   \\
 1.01289 & 0.33275 & 0.3376 & 0.02997 & 1.41188 & 2.49881 & 0.50628 & 0.00183   \\
 1.10247 & 0.33371 & 0.3419 & 0.02939 & 1.43774 & 2.53384 & 0.44079 & 0.00195   \\
\hline                                                                  
1.03009 & 0.33228 & 0.33905 & 0.02902 & 1.44506 & 2.50031 & 0.46076 & 0.00165 \\
\hline
\end{tabular}
\caption{Estimated Parameters on 11 different runs with low noise and no signal present.}
\label{tab:NoiseOnlyResults} 
\end{table}

@@

\subsection{Single Voxel Review}
\begin{table}[t]
\centering
\begin{tabular}{|c | c | c | c | c | c | c | c | c |}
\hline 
& \multicolumn{4}{|c|}{Signal} & \multicolumn{4}{|c|}{No Signal}\\
\hline
& \multicolumn{2}{|c|}{Low Noise} & \multicolumn{2}{|c|}{High Noise} 
& \multicolumn{2}{|c|}{$\sigma_y = .001, \sigma_x = .0005$} 
& \multicolumn{2}{|c|}{$\sigma_y = .01, \sigma_x = .005$}\\
\hline 
& $M.I.$ & N. Res. &
  $M.I.$ & N. Res. &
  $M.I.$ & N. Res. &
  $M.I.$ & N. Res. \\
\hline       
\hline       
1 &   0.86687  &0.47801 &  0.09077  &1.03894 &  0.06326  &  1.29501 & 0.03024  &1.33641 \\
2 &   0.93975  &0.53177 &  0.13767  &0.95165 &  -0.01075  & 1.30175 & -0.02677 &1.33667 \\
3 &   0.82382  &0.5458  &  0.13505  &0.99539 &  0.02345  &  1.26287 & -0.0111  &1.15957 \\
4 &   0.94661  &0.49824 &  0.04341  &1.16129 &  -0.00906  & 1.43196 & 0.00147  &1.09988 \\
5 &   0.94281  &0.46805 &  0.13718  &1.03972 &  0.00663  &  1.25664 & -0.00204 &1.20107 \\
6 &   0.92539  &0.459   &  0.12337  &1.00214 &  -0.00816  & 1.2708 &  0.01775  &1.04589 \\
7 &   0.98892  &0.46096 &  0.15381  &1.08847 &  0.02664  &  1.15441 & 0.03163  &1.20543 \\
8 &   0.98796  &0.51838 &  0.11325  &1.05962 &  0.03285  &  1.27456 & 0.01951  &1.1225 \\
9 &   0.8804   &0.5253  &  0.09669  &1.0157  &  0.01628  &  1.32024 & 0.01039  &1.08637 \\
10 &  0.88721   &0.49211 & 0.18339  &1.18996 &  0.00407  &  1.34456 & 0.00508  &1.22135 \\
11 &  0.96644   &0.49092 & 0.10949  &0.95368 &  0.03323  &  1.32522 & -0.01284 &1.11737 \\
\hline                                                                         
mean &0.92329  &0.49714 & 0.12037   &1.04514 &  0.01622  &  1.29437 & 0.00576  &1.17568 \\
\hline                                                                           
min &  0.82382  &0.459   &  0.04341  &0.95165 & -0.01075  & 1.15441 & -0.02677 &1.04589 \\
\hline                                                                         
max &  0.98892  &0.5458  & 0.18339   &1.18996 & 0.06326  &  1.43196 & 0.03163  &1.33667 \\
\hline
\end{tabular}
\caption{Mutual Information and the normalized $\sqrt{MSE}$, 
    for signal/noise configurations.}
\label{tab:SignleVoxelActivationComparison} 
\end{table}

The first step in determining the validity of a model is to provide some order of quality
to rate the results by. Unfortunately, because of the variability in the signal levels the
raw $\sqrt{MSR}$ cannot perform this task. As demonstrated by \autoref{sec:PureNoiseLowMag},
a low residual does not necessarily indicate a good fit. 
Therefore, instead I  normalized the signal based on a value proportional to
the magnitude of the signal. Considering the tendency of FMRI noise to have large unexplainable
peaks and troughs, rather than using MAX and MIN values to estimate the scale of the 
signal, I used a robust estimator of 
scale; the Median-Absolute-Deviation (MAD). This is an estimator of the standard
deviation, and thus a good estimator of the scale of the input signal. 
The normalized residual values are 
shown in \autoref{tab:SignleVoxelActivationComparison}. A second potential method of 
gauging performance is mutual information. Mutual information is a method of measuring
the interdependence of two random variables. If two signals are truly independent,
then the mutual information will be zero. Although ideally suited to Bernoulli 
random variables where there are a finite number of options, by using histograms
it is possible to derive a joint distribution of two signals. The algorithm for 
mutual information is based on that joint distribution:
\begin{equation}
\sum_{x,y} p(x,y) \log_2\left(\frac{p(x,y)}{p(x)p(y)}\right)
\end{equation}
Unfortunately the number of bins causing bias in the output, thus to correct for this,
I subtracted the estimated bias:
\begin{equation}
\text{bias} = \frac{N_{bins}}{2Nlog(2)}
\end{equation}
where $N$ is the number of samples and $N_{bins}$ is the number of bins. For all
the mutual information estimates in this work 6 bins are used for the marginal
distribution of each signal. Additionally, throughout log base 2 will be used.
This leads to 36 total bins in the joint, so the bias is: 
\begin{equation}
\text{bias} = \frac{18}{N}
\end{equation}
After calculating the empirical mutual information, I subtracted the bias from
the estimated mutual information. Note that this can result in negative mutual
information, which should not technically be possible; so any negative mutual information
was taken as 0.

@@

\begin{figure} %bottom left
\centering
\includegraphics[clip=true,trim=2.5cm 2cm 2cm 1cm,width=15cm]{images/slicesim_hist1}
\caption{Histogram of estimated parameters in section 1 in voxels with mutual information greater
than $.14$}
\label{fig:slicesim_hist1}
\end{figure}

\begin{figure} %top left
\centering
\includegraphics[clip=true,trim=2.5cm 2cm 2cm 1cm,width=15cm]{images/slicesim_hist2}
\caption{Histogram of estimated parameters in section 2 in voxels with mutual information greater
than $.14$}
\label{fig:slicesim_hist2}
\end{figure}

\begin{figure} %top right
\centering
\includegraphics[clip=true,trim=2.5cm 2cm 2cm 1cm,width=15cm]{images/slicesim_hist3}
\caption{Histogram of estimated parameters in section 3 in voxels with mutual information greater
than $.14$}
\label{fig:slicesim_hist3}
\end{figure}
todo regenerate histograms

### 2010-07-22 doc/thesis/resultsB.tex
fixed up results.
fixed the histograms for slicesim

In reality, the statistical parametric mapping method is very different from the 
particle filter algorithm described in this work.
SPM preprocesses the image by spatially smoothing the FMRI image (in this section 
SPM8 was used with an $8mm x 8mm x 8mm$ Gaussian kernel), whereas
this is not done in the particle filter algorithm. Additionally, a spline
was used to de-trend, rather than SPM8's high pass filter (with a cut
off based on a globally estimated autocorrelation). Thus the preprocessing pipelines 
are different; but the output of SPM8 is also different. Whereas SPM outputs
a t-statistic for each voxel, the output 
of the particle filter is a posterior probability distribution of the parameters
at every voxel. To validate the quality of the particle filter results though
it is necessary to compare the location and fit of the particle filter parameter
mapping with SPM's T-score map.

@@

The results from SPM8 are shown in \autoref{fig:hm_canon_spm}, and the results from 
the particle filter are shown in \autoref{fig:hm_canon_pfilter}. Note again that the
output is actually a different scale. Whereas the levels in the SPM map are T-Scores
indicating the probability that the result cannot be explained from a random signal,
the result of the particle filter is simply a normalized square-root of the mean-squared-
residual. Thus, lower is better, and zero is a perfect fit in the particle filter whereas
infinity is a perfect fit in \autoref{fig:hm_canon_spm}. Note that it would seem that there
are quite a few dispersed areas with decent fits in \autoref{fig:hm_canon_pfilter}. Also,
the lowest error was just above $.7$. 
For the sake of completeness \autoref{fig:hm_canon_pfilter85} shows the same image but
thresheld at $.85$ rather than $1$. Obviously this image seems like a closer match to
the results of SPM, however, the results of \autoref{fig:hm_canon_pfilter} are interesting.

@@

\begin{figure}[H]
\subfigure[]{\label{fig:hm_mi} \includegraphics[scale=.85]{images/mi_hm}}
\subfigure[]{\label{fig:hm_canon_mi_x} \includegraphics[scale=3]{images/mi_hm_x}}
\subfigure[]{\label{fig:hm_canon_mi_y} \includegraphics[scale=3]{images/mi_hm_y}}
\subfigure[]{\label{fig:hm_canon_mi_z} \includegraphics[scale=3]{images/mi_hm_z}}
\subfigure{\label{fig:scale_mi} \includegraphics[scale=.5]{images/scale4}}
\caption{Sagittal, coronal and axial (\autoref{fig:hm_canon_pfilter_x} \autoref{fig:hm_canon_pfilter_y} 
         \autoref{fig:hm_canon_pfilter_x}), as well as a series of axial slices, \autoref{fig:hm_pfilter}. 
         Units of match is bits (standard for base-2 Mutual Information). The highest (best) levels are 
         $.42$. The worst shown is $.1$.}
\label{fig:hm_canon_pfilter85}
\end{figure}

>>>>

In reality, the purpose of statistical parametric mapping method is 
very different from the 
particle filter algorithm described in this work.
SPM preprocesses the image by spatially smoothing the FMRI image (in this section 
SPM8 was used with an $8mm x 8mm x 8mm$ Gaussian kernel), whereas
this is not done in the particle filter algorithm. Additionally, a spline
was used to de-trend, rather than SPM8's high pass filter (with a cut
off based on a globally estimated autocorrelation). Thus the preprocessing pipelines 
are different; but the output of SPM8 is also different. Whereas SPM outputs
a t-statistic for each voxel, the output 
of the particle filter is a posterior probability distribution of the parameters
at every voxel. To validate the quality of the particle filter results though
it is necessary to compare the location and fit of the particle filter parameter
mapping with SPM's T-score map.

@@

The results from T-values from SPM8 are shown in \autoref{fig:hm_canon_spm} (threshold of
4), and the results from 
the particle filter are shown in \autoref{fig:hm_canon_pfilter} and \autoref{fig:hm_canon_pfilter_mi}.
Note that the scales for all three images are different, because the metrics are different.
SPM measures using T-Tests to determine the likelihood of a false positive. \autoref{fig:hm_canon_pfilter}
uses simple normalized residuals, meaning that lower indicates less error.  
\autoref{fig:hm_canon_pfilter_mi} measures in terms of the dependence between the 
measured signal and the estimated signal; thus higher indicates a better fit. The particle
filter data shows a large number of false positives, however application of a threshold
of $.85$ on the residual map removes these false positives. Similarly, in the mutual information
map, the false positives may be elminated by upping the threshold to $.15$. However, just
because the results disagree with SPM does not necessarily mean they are false positives.
SPM operates on smoothed data (8mm x 8mm x 8mm), so there are certainly active 
areas that have been missed because of the smoothing.

@@

\begin{figure}[H]
\subfigure[]{\label{fig:hm_mi} \includegraphics[scale=.85]{images/mi_hm}}
\subfigure[]{\label{fig:hm_canon_mi_x} \includegraphics[scale=3]{images/mi_hm_x}}
\subfigure[]{\label{fig:hm_canon_mi_y} \includegraphics[scale=3]{images/mi_hm_y}}
\subfigure[]{\label{fig:hm_canon_mi_z} \includegraphics[scale=3]{images/mi_hm_z}}
\subfigure{\label{fig:scale_mi} \includegraphics[scale=.5]{images/scale4}}
\caption{Sagittal, coronal and axial (\autoref{fig:hm_canon_pfilter_x} \autoref{fig:hm_canon_pfilter_y} 
         \autoref{fig:hm_canon_pfilter_x}), as well as a series of axial slices, \autoref{fig:hm_pfilter}. 
         Units of match is bits (standard for base-2 Mutual Information). The highest (best) levels are 
         $.42$. The worst shown is $.1$.}
\label{fig:hm_canon_pfilter_mi}
\end{figure}
